Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-22 paper.tex
FINAL VERSION
related work

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Ond\v{r}ej V\'{a}vra$^{3}$,
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic, {\tt vonasek@labe.felk.cvut.cz};
$^{2}$ Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic;
$^{3}$ Loschmidt Laboratories, Department of Experimental Biology RECETOX, Faculty of Science, Masaryk University
    Kamenice 5, 625 00 Brno, Czech Republic. 
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
}}
%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%}%
%}
%}

@@

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{tunnel}}
\def\da{d_{atom}}
\def\R{\mathbb{R}}
\def\QI{Q_{init}}
\def\RI{R_{init}}

@@

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere.
This can lead to highly imprecise results.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
This can be solved using sampling-based planning methods.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations, which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

\section{Related Work}

Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing the protein-ligand interaction possibilities.
More details can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
The ligand flexibility cannot be principally considered by these approaches.
Simulations based on molecular dynamics (MD) can directly evaluate the migration of the ligand into the protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of different ligands.
The geometry-based methods are therefore still preferable due to their speed.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF, that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using the electrostatic energy.
The high-energy configurations are discarded, while the low-energy conformations (and their vicinity) are sampled densely.
The approach~\cite{singhLIG} can be also used to predict the ligand binding sites.

@@

While the classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures the distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.

@@

The proposed algorithm is based on the principle of the RRT~\cite{lavalleRRT} planner.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
both 3D rotation and 3D translation.
The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following three proposed modifications.

@@

\linesnumbered
\begin{algorithm}[hb]
{\small
\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$,
   distance $\dt$ to move virtual goal
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
$\T$.addNode($\qinit$)\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\T, \qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

@@

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
the retraction of the ligands towards $\qrand$.

@@

The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest atom from $q$ towards atoms of $\qrand'$.
%is also the nearest point of the ligand at $q$ to $\qrand$.

\begin{algorithm}[th]
{\small
\setstretch{0.88}
\caption{\label{alg::expand}expand}
\KwIn{
   tree $\T$,
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \nllabel{alg::expb}
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to the next conformation
        }
    }
}
}
\end{algorithm}

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
The \RA\ method was also tested without the atom-based metric.
In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.

@@

Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
the planner prefers to apply the higher scales if possible.
%The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
%that leads to a new collision-free configuration.
The information about the scales used at the trajectory points 
brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
According to the results of the motion planing, the DCP ligand needs to shrink down to ca. 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
Beyond this point, the ligand can move at larger scales again.
As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
%\includegraphics[width=0.2\textwidth]{fig/colormap}
\caption{\label{fig::dcp}
    \small
Visualization of trajectories for the ligand DCP in the two examined tunnels.  
The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to be scaled-down (to $\sim75$~\%) to pass the
next part of the tunnel (B). After this narrow passage, the tunnel is larger again and the ligand can traverse to the narrow passage with
scale $>75$~\% (C).
The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
The ligand can reach the active site only at a very low scale ($\sim0.4=40$~\%) using the side tunnel.
}
\end{figure}

>>>>

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Ond\v{r}ej V\'{a}vra$^{3}$
%    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic, {\tt vonasek@labe.felk.cvut.cz};
$^{2}$ Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic;
$^{3}$ Loschmidt Laboratories, Department of Experimental Biology RECETOX, Faculty of Science, Masaryk University
    Kamenice 5, 625 00 Brno, Czech Republic. 
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
}}
%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%}%
%}
%}

@@

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{tunnel}}
\def\da{d_{atom}}
\def\dist{d}
\def\R{\mathbb{R}}
\def\QI{Q_{init}}
\def\RI{R_{init}}

@@

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to highly imprecise results.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations, which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

\section{Related Work}
{\color{red}
Migration of the ligands into a protein can be directly evaluated using Molecular dynamics (MD) simulations, that consider both 
the protein and the ligand flexible.
However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
for particular ligand/protein pair, which is not practical for testing a large number of different ligands.
Therefore, computational approaches that consider only geometry of the protein/ligand are still preferable due to their high speed.

The widely used approaches to evaluate possiblity of ligand migration into the protein rely on the tunnel detection.
Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing the protein-ligand interaction possibilities.
The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations.
The protein conformations can be prepared again using MD simulations.
Computationally less costly approaches for generating protein conformations are based on constrained geometric simulations like
ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained} and tCONCOORD~\cite{seeliger2007geometry}.
Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.

The main disadvantage of both grid-based and VD-based tunnel detection methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
The ligand flexibility cannot be principally considered by Voronoi-based approaches.
}

%Simulations based on molecular dynamics (MD) can directly evaluate the migration of the ligand into the protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
%Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of different ligands.
%The geometry-based methods are therefore still preferable due to their speed.

@@

%The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot or the obstacles~\cite{bayazitIRC}.

The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF, that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
Each configuration in the roadmap is also evaluated using the electrostatic energy.
The high-energy configurations are discarded, while the low-energy conformations (and their vicinity) are sampled densely.
The approach~\cite{singhLIG} can also predict the ligand binding sites.

@@

%While the classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
%This metric measures the distances between two configurations as the nearest distance of two atoms. 
%This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.

@@

The proposed algorithm is based on the principle of the RRT~\cite{lavalleRRT} planner.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
both 3D rotation and 3D translation.
The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following two proposed modifications.

@@

\linesnumbered
\begin{algorithm}[b]
{\small
\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$,
   distance $\dt$ to move virtual goal
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
$\T$.addNode($\qinit$)\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\T, \qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

@@

%The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
%This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
%To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
%The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
%$\da(q_1,q_2)$.
%This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
%It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
%the retraction of the ligands towards $\qrand$.

@@

%The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
%For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
%(i.e., for ligand with no translation and no rotation).
%To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
%rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
%Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest atom from $q$ towards atoms of $\qrand'$.
%

\begin{algorithm}[b]
{\small
\setstretch{0.88}
\caption{\label{alg::expand}expand}
\KwIn{
   tree $\T$,
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\dist(q, \qrand) < \dist(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \nllabel{alg::expb}
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to the next conformation
        }
    }
}
}
\end{algorithm}

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
%The \RA\ method was also tested without the atom-based metric.
%In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.

@@

Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
the planner prefers to apply the higher scales if possible.
%The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
%that leads to a new collision-free configuration.

The information about the scales used at the trajectory points 
brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
According to the results of the motion planing, the DCP ligand needs to shrink down to ca. 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
Beyond this point, the ligand can move at larger scales again.
As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.

@@

\begin{figure}
\centering
\includegraphics[width=0.35\textwidth]{fig/dcp-image1Label}\\
%\includegraphics[width=0.2\textwidth]{fig/colormap}
\vspace{-3mm}
\caption{\label{fig::dcp}
    \small
Visualization of trajectories for the ligand DCP in the two examined tunnels.  
The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to be scaled-down (to $\sim75$~\%) to pass the
next part of the tunnel (B). After this narrow passage, the tunnel is larger again and the ligand can traverse to the narrow passage with
scale $>75$~\% (C).
The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
The ligand can reach the active site only at a very low scale ($\sim0.4=40$~\%) using the side tunnel.
}
\end{figure}
